Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-06-24 Text/migration.tex
Manuscript prepared for resubmission to ApJL
Added h/r plot and text to address several of the referee's concerns.
	modified:   ../README.md
	modified:   migration.tex
	modified:   migration_apjlsubmission.tex

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, is simulated from time $t = 0$
to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass
($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25
AU. In the third simulation, which we call the {\sl t = 10 planet run}, a 1 $M_J$ planet with a similar orbit is
inserted into the fiducial run at $t = 10.5$ ORP. These three simulations are part of a larger suite, to be described
elsewhere, in which the planet mass is varied. We estimate semi-major axes $a$ and eccentricities $e$ from the planet's
$\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, over each complete orbit
of the planet, where one orbit is $\Delta\phi=2\pi$. Specifically, $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e =
(\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These parameters are intended to highlight the
strong orbit-to-orbit variations in the motion of the planet.  A more refined estimate of the orbital elements is
unnecessary because the potential of the disk, even if were axisymmetric, will cause departures from Keplerian dynamics.

@@

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too
different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3
ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae
that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from
the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation. While the interaction with a
GI-active disk does not seem to result in a drastically different overall migration rate, the direction and magnitude of
migration in both simulations is quite variable and seems to depend strongly on interactions of the planet with discrete
modes in the disk. In $t = 10$ planet run, the overall radial migration rate is similar but is even more erratic in
direction than in the $t = 0$ planet run. The important departures from laminar disk theory are threefold: the more
erratic nature of migration, the importance of interactions with discrete low-order modes, and stalling of migration
near the inner edge of the strongest GI activity. The migration is neither a random walk nor monotonic but has a chaotic
character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the
dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$
type runs) and relative to the CRs of low-order GI modes (for $t = 10$ type runs).

@@

Due to the global nature of the GIs and their strong interaction with the embedded planet, it is important that the gas
phase physics are as accurate as possible and detailed analysis of the interaction between the planet and global modes
is required. Of particular importance is the treatment of cooling in the disk. Although these simulations show some
similarities to \citet{baruteau2011}, their simulations employ an idealized cooling rate governed by the local dynamical
timescale while the simulations in this work use the more realistic radiative physics of \citet{boley2007b}. {\bf
  citations on global vs. local needed} Another primary difference between this work and that of \citet{baruteau2011} is
the simulations presented here are three dimensional. Many studies have shown that three dimensional simulations are
necessary to accurately measure the effects of GIs {\bf need some more citations here}. Further in depth analysis of the
GIs and the interaction between the GIs and the planet will be presented in a future work.

{\bf Say something about possible physical scenarios for t=0 run}

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research
was conducted with the support of NASA grants from the Origins of Solar Systems Program (NNG05GN11G and
NNX08AK36G). S.M. was supported during portions of this research by an NESSP Graduate Fellowship and by Indiana Space
Grant Consortium Graduate Fellowships; A.C.B. is currently a Sagan Postdoctoral Fellow at U. Florida. This material is
based upon work supported by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258,
OCI-0504075, and CNS-0521433, and through TeraGrid resources provided by Indiana University, PSC, SDSC, and
NCSA. TeraGrid systems are hosted by Indiana University, LONI, NCAR, NCSA, NICS, ORNL, PSC, Purdue University, SDSC,
TACC, and UC/ANL. Any opinions, findings and conclusions or recommendations expressed in this material are those of the
authors and do not necessarily reflect the views of the National Science Foundation.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/hr_migration.eps}
\caption{(Color figure available in the online version) Plots of the disk aspect ratio $h/r$ as determined by a decrease
  from midplane density by a factor of $e^2$ for the fiducial and $t=0$ runs (black) at $\approx 10$ ORP and the fiducial and
  $t=10$ runs (red) at $\approx 20$ ORP.}
\label{fig:hr}
\end{figure}

@@

>>>>

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, is simulated from time $t = 0$
to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass
($M_J$) planet with a roughly circular midplane orbit at 25 AU is included in the initial $t = 0$ ORP equilibrium
disk. This run investigates the effects of an existing planet on the growth of GIs, which could occur if disk the disk
undergoes episodic or localized instabilities. In the third simulation, which we call the {\sl t = 10 planet run}, a 1
$M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. We estimate semi-major axes $a$
and eccentricities $e$ from the planet's $\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and
$\varpi_{\rm min}$, over each complete orbit of the planet, where one orbit is $\Delta\phi=2\pi$. Specifically, $a =
(\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm
  min})$. These parameters are intended to highlight the strong orbit-to-orbit variations in the motion of the planet.
A more refined estimate of the orbital elements is unnecessary because the potential of the disk, even if were
axisymmetric, will cause departures from Keplerian dynamics.

@@

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too
different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3
ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae
that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from
the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation (see figure \ref{fig:hr} for
disk aspect ratios). While the interaction with a GI-active disk does not seem to result in a drastically different
overall migration rate, the direction and magnitude of migration in both simulations is quite variable and seems to
depend strongly on interactions of the planet with discrete modes in the disk. In $t = 10$ planet run, the overall
radial migration rate is similar but is even more erratic in direction than in the $t = 0$ planet run. The important
departures from laminar disk theory are threefold: the more erratic nature of migration, the importance of interactions
with discrete low-order modes, and stalling of migration near the inner edge of the strongest GI activity. The migration
is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons,
it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its
placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes
(for $t = 10$ type runs).

@@

Due to the global nature of the GIs and their strong interaction with the embedded planet, it is important that the gas
phase physics are as accurate as possible and detailed analysis of the interaction between the planet and global modes
is required. Of particular importance is the treatment of cooling in the disk. Although these simulations show some
similarities to \citet{baruteau2011}, their simulations employ an idealized cooling rate governed by the local dynamical
timescale while the simulations in this work use the more realistic radiative physics of
\citet{boley2007b}. Three-dimensional simulations with radiative physics are necessary for understanding the evolution
of spiral structure in vertically stratified disks because shocks and heating along spiral arms can change the disk's
vertical structure \citep[e.g.][]{boley2006a}, which affects the migration rate of planets.

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research
was conducted with the support of NASA grants from the Origins of Solar Systems Program (NNG05GN11G and
NNX08AK36G). A.C.B. is currently a Sagan Postdoctoral Fellow at U. Florida. This material is based upon work supported
by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258, OCI-0504075, and
CNS-0521433, and through TeraGrid resources provided by Indiana University, PSC, SDSC, and NCSA.

@@

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/hr_migration.eps}
\caption{(Color figure available in the online version) Plots of the disk aspect ratio $h/r$ as determined by a decrease
  from midplane density by a factor of $e^2$ for the fiducial and $t=0$ runs (black) at $10.5$ ORP and the fiducial and
  $t=10$ runs (light gray) at $20.5$ ORP.}
\label{fig:hr}
\end{figure}

### 2011-06-24 Text/migration_apjlsubmission.tex
Final, Final, Final edits. Added Dick's comment and submitted 06/07/2011
	modified:   Text/migration.tex
	modified:   Text/migration_apjlsubmission.tex

git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@87 3b7f7548-5c66-0410-83ab-3997167635fc
Manuscript prepared for resubmission to ApJL
Added h/r plot and text to address several of the referee's concerns.
	modified:   ../README.md
	modified:   migration.tex
	modified:   migration_apjlsubmission.tex

The model disk, based on an equlibrium model from \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007} minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed nonaxisymmetry, the density distribution is given an initial 0.01 \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the {\sl fiducial run},  is simulated from time $t = 0$ to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the {\sl t = 10 planet run}, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. These three simulations are part of a larger suite, to be described elsewhere, in which the planet mass is varied. We estimate semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, over each complete orbit of the planet, where one orbit is $\Delta\phi=2\pi$. Specifically, $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These parameters are intended to highlight the strong orbit-to-orbit variations in the motion of the planet.  A more refined estimate of the orbital elements is unnecessary because the potential of the disk, even if were axisymmetric, will cause departures from Keplerian dynamics.

@@

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation. This is a reasonable estimate for comparison purposes, because our resolution is probably insufficient to compute the corotation torque \citep{paardekooper2010}. 
While the interaction with a GI-active disk does not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in both simulations is quite variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In $t = 10$ planet run, the overall radial migration rate is similar but is even more erratic in direction than in the $t = 0$ planet run. The important departures from laminar disk theory are threefold: the more erratic nature of migration, the importance of interactions with discrete low-order modes, and stalling of migration near the inner edge of the strongest GI activity. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes (for $t = 10$ type runs).

@@

@@

\begin{figure}[t]
\center
\includegraphics[width=11cm]{amt.eps}
\caption{(Color figure available in the online version) Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The top and bottom panels correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively.}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{planeta.eps}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The dashed-diamond and solid-star lines (colored red and black in the online version) indicate the curves ($\varpi(t)$), points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The symbols are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{planete.eps}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text, as a function of time. The diamonds and stars (colored red and black in the online version) indicate the points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}

>>>>

The model disk, based on an equilibrium model from \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d =
0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim
\varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk
at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is
close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007}
minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to
accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the
upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed
nonaxisymmetry, the density distribution is given an initial 0.01 \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, is simulated from time $t = 0$
to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass
($M_J$) planet with a roughly circular midplane orbit at 25 AU is included in the initial $t = 0$ ORP equilibrium
disk. This run investigates the effects of an existing planet on the growth of GIs, which could occur if disk the disk
undergoes episodic or localized instabilities. In the third simulation, which we call the {\sl t = 10 planet run}, a 1
$M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. We estimate semi-major axes $a$
and eccentricities $e$ from the planet's $\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and
$\varpi_{\rm min}$, over each complete orbit of the planet, where one orbit is $\Delta\phi=2\pi$. Specifically, $a =
(\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm
  min})$. These parameters are intended to highlight the strong orbit-to-orbit variations in the motion of the planet.
A more refined estimate of the orbital elements is unnecessary because the potential of the disk, even if were
axisymmetric, will cause departures from Keplerian dynamics.

@@

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too
different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3
ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae
that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from
the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation (see figure \ref{fig:hr} for
disk aspect ratios). While the interaction with a GI-active disk does not seem to result in a drastically different
overall migration rate, the direction and magnitude of migration in both simulations is quite variable and seems to
depend strongly on interactions of the planet with discrete modes in the disk. In $t = 10$ planet run, the overall
radial migration rate is similar but is even more erratic in direction than in the $t = 0$ planet run. The important
departures from laminar disk theory are threefold: the more erratic nature of migration, the importance of interactions
with discrete low-order modes, and stalling of migration near the inner edge of the strongest GI activity. The migration
is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons,
it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its
placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes
(for $t = 10$ type runs).

@@

Due to the global nature of the GIs and their strong interaction with the embedded planet, it is important that the gas
phase physics are as accurate as possible and detailed analysis of the interaction between the planet and global modes
is required. Of particular importance is the treatment of cooling in the disk. Although these simulations show some
similarities to \citet{baruteau2011}, their simulations employ an idealized cooling rate governed by the local dynamical
timescale while the simulations in this work use the more realistic radiative physics of
\citet{boley2007b}. Three-dimensional simulations with radiative physics are necessary for understanding the evolution
of spiral structure in vertically stratified disks because shocks and heating along spiral arms can change the disk's
vertical structure \citep[e.g.][]{boley2006a}, which affects the migration rate of planets.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/amt.eps}
\caption{(Color figure available in the online version) Global amplitudes of nonaxisymmetric density perturbations $A_m$
  as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of
  \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge
  effects in the inner disk. The top and bottom panels correspond to the fiducial run (no planet) and the $t = 0$ planet
  run, respectively.}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planeta.eps}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The
  dashed-diamond and solid-star lines (colored red and black in the online version) indicate the curves ($\varpi(t)$),
  points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The
  symbols are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as
  explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planete.eps}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in
  the text, as a function of time. The diamonds and stars (colored red and black in the online version) indicate the
  points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/hr_migration.eps}
\caption{(Color figure available in the online version) Plots of the disk aspect ratio $h/r$ as determined by a decrease
  from midplane density by a factor of $e^2$ for the fiducial and $t=0$ runs (black) at $10.5$ ORP and the fiducial and
  $t=10$ runs (light gray) at $20.5$ ORP.}
\label{fig:hr}
\end{figure}
